---
title: "Regular Expressions R-Club"
output: html_document
author: Original docment by Matt Jones at NCEAS, edited by Alex Ilich
editor_options: 
  chunk_output_type: inline
---

```{r,echo=FALSE, include=FALSE}
options(stringsAsFactors = FALSE)
# Fake data generation code
set.seed(5)
site_data <- data.frame(year = rep(seq(2000, 2009), 4))
site_data$site <- sample(c("galveston bay", "choctawhatchee bay", "aransas bay", "copany bay"), nrow(site_data), replace = TRUE)
site_data$subsite <- sample(c(1,2,3,4), nrow(site_data), replace=TRUE)
site_data$temp_c <- runif(nrow(site_data), 0, 100)
site_data$site_code <- paste(site_data$year, site_data$site, site_data$subsite, sep="-")
site_data <- site_data[,c("site_code", "temp_c")]
names(site_data) <- c("site_code", "temp_c")
```

# Introduction

Regular expressions are a fantastic tool for filtering and even extracting information out of strings of characters such as site codes, titles, or even entire documents. 
Regular expressions follow a custom syntax that we'll need to learn but they are worth learning because:

- Regular expressions can do things other methods cannot
- Regular expressions can be used with many other languages and tools so it's a learn-once, use-everywhere kind of tool

But they're something that you only need to learn a bit of to get a lot of value out of them.

# Learning Outcomes

- Understand when regular expressions are appropriate
- Have an introductory-level awareness of regular expression syntax
- Have some experience executing and working with regular expressions in R

# Lesson

The `stringr` package from the `tidyverse` is useful for working with regular expressions in R
```{r warning=FALSE, message=FALSE}
library(stringr)
library(dplyr)
```

Let's start off with a simple example of where simpler methods won't work and see how regular expressions can be used to get what we need done.
Let's say we just received some data we need to analyze and we find this:

```{r}
site_data
```


It looks like the author of the dataset mixed the year of measurements, site code (e.g., A, CCCC, etc.), and some sub-site code (e.g., 1, 2, 3, etc.) into a single column.
If we wanted to, for example, calculate mean temperature by site, we'd need to split these up somehow into separate columns.
How could we go about this?
We could start with `substr` which lets us slice a string by its indices:

```{r}
substr(site_data$site_code, 1, 4)
substr(site_data$site_code, 5, 16)
```

But we'd quickly find that, because the number of characters in the site code varies from one to four, we can't extract just the site code. 
These are the types of problems where regular expressions come in handy.

Before we start, we're going to use the `str_view_all` function from the `stringr` package which shows a nice display of the result of executing a regular expression against our strings.
In real use, we would use another function to actually get and work with the result.

```{r message=FALSE}
str_view_all(string = site_data$site_code, pattern = "[a-z ]+")
site_data %>% mutate(site=str_extract(site_code, pattern = "[a-z ]+"))
```

The expression we used above, `[a-z ]+`, is equivalent to asking for the first consecutive run of the letters a-z or a space in the entire string of characters.
This is the type of problem regular expression were created for!


## Overview of Regular Expressions

### Literal Matches
Regular expressions can match things literally, e.g.,

```{r}
str_detect(string = "red grouper", pattern = "snapper")
str_detect(string = "red grouper", pattern = "grouper")
```

but they also support a large set of special characters:

### Special Characters
There are also many special characters we can use  

- `.` indicates any character  

- `+` matches one or more times  

- `*` matches 0 or more times  

- `{n}` matches exactly n times  

- `{n,}` matches at least n times  

- `{,n}` matches at most n times  

- `()` groupings  

- `^` start of string  

- `$` end of string  

- `\\<` Beginning of a word  

- `\\>` End of a word  

- `[]`: Match any character in this set  

- `[^]`: Match any character *not* in this set  

- `\\d` digits  

- `\\D` non-digits  

- `\\s` whitespace  

- `\\S` non-whitespace  

- `\\w` any word character [A-Za-z0-9_]  

- `\\W` any non-word character  

- `?` optional  

- `\\` escape used to indicate you are searching for one of the special characters  

### Common R functions that use regular expressions
- Base R  
  - `grep`/`grepl` Find indices (numeric/logical) of vector elements that match the specified pattern.  
  - `sub`/`gsub` Replace a pattern with something else  
  - `strsplit` Split the elements of a character vector x into substrings according to the matches to substring split within them.  
- `stringr` package  
  - `str_detect` Find indices (logical) of vector elements that match the specified pattern. Works like grepl.  
  - `str_match`/`str_match_all` Extract matched groups from a string  
  - `str_replace`/`str_replace_all` Replace a pattern with something else (like sub/gsub)  
  - `str_split`/`str_split_fixed` Split up a string into pieces (like strsplit)  
  - `str_locate`/`str_locate_all` Finds start and end character position of a pattern within a string  
  - `str_extract`/`str_extract_all` Extract pattern  
  - `str_remove/str_remove_all` Remove pattern  
  
Notes:  
  
  - Base R functions have arguments in order of pattern, string whereas stringr functions have arguments in the order of string, pattern because this allows for piping.  
  
  - If multiple matches are found, many functions will just use the first match. In stringr the functions will the suffix "all" will act on all matches.  

### Examples

- The period (`.`) matches any character

```{r}
fish <- c("gag grouper", "red snapper", "sheepshead")
str_view_all(fish, ".p")
```

- `[]`: Match any character in this set

```{r}
DNA<- c("ATGTTAGCTTAGCCTA", "ATGATGCTCA", "TTCGGTTAGCAG")
target_seq<- "ATG[CT]"
str_view_all(DNA, target_seq)
str_detect(DNA, target_seq)
grepl(target_seq,DNA)
grep(target_seq,DNA)
grep(target_seq,DNA, value=TRUE)
str_locate(DNA, target_seq)
```


- `[^]`: Match any character *not* in this set

```{r}
str_view_all(fish, "[^aeiou]")
```

- `\\s` to match any whitespace (e.g. Space, tab, vertical tab, newline) & `\\S` to match any non-whitespace

```{r}
str_view_all(fish, "\\s")
str_view_all(fish, "\\S")
```

- `\\d` to match any digit which is equivalent to `[0-9]` & `\\D` to match any non-digit

```{r}
str_view_all(site_data$site_code, "\\d")
str_view_all(site_data$site_code, "\\D")
```

- `\\w` to match any word character `[A-Za-z0-9_]` & `\\W` to match any non-word character

```{r}
str_view_all(site_data$site_code, "\\w")
str_view_all(site_data$site_code, "\\W")
```


- `?` to specify if a character/group of characters is optional  

- `{n}` to specify how many times a character/group of characters will occur  

Say we want to get just the phone numbers out of this vector but we notice that the phone numbers take on some different formats:

```{r}
contact_info<- data.frame(form_submission= c("219 733 8965", "user@usf.edu", "329-293-8753 ", "governmentemployee1@fwc.gov", "595.794.7569", "3872876718"))
contact_info
str_view_all(string = contact_info$form_submission, pattern = "\\d{3}[ \\.-]?\\d{3}[ \\.-]?\\d{4}")
```
The above regular expression matches the number parts of the phone numbers, which can be separated by zero or one space, `.`, or `-`.

We could use this for example to clean responses into a common format including only digits
```{r}
phone_numbers<- contact_info %>% filter(str_detect(string = form_submission,pattern = "\\d{3}[ \\.-]?\\d{3}[ \\.-]?\\d{4}"))
phone_numbers %>% mutate(formatted_num= str_remove_all(string = form_submission, pattern = "\\D")) #Remove all non-digits
```

We can use the `+` expression to find sequences of at least 1 of a match

```{r}
str_view_all(site_data$site_code, "\\d+")
```

- The `*` is zero or more matches

By default the asterisk (`*`) is greedy, i.e. it always matches the longest possible string. 
It can be used in lazy mode by adding?, i.e. `*?`.

```{r}
letters <- "abcdefghijkb"
str_view_all(letters, "a.*b") # Greedy
str_view_all(letters, "a.*?b") # Lazy
```


- `()`: Grouping

One of the most powerful parts of regular expressions is grouping.
Grouping allows us to split up our matched expressions and do more work with them.
For example, we can create match the city and state in a set of addresses, splitting it into components:

```{r}
addresses <- c("Santa Barbara, CA", "Seattle, WA", "New York, NY")
str_match(addresses, "([\\w\\s]+), (\\w+)")
```
In `str_match`, the first column is the complete match, followed by one column for each capture group. 

Once we use groups, `()`, we can also use back references to work with the result.
Back references are `\\` and a number, where `\\1` is the first thing in `()`, `\\2` is the second thing in `()`, and so on.

```{r}
addresses
str_replace(string = addresses, pattern= "([\\w\\s]+), (\\w+)", "City: \\1, State: \\2")
```

- `|` for or
```{r}
fish2<- c("red grouper", "red snapper", "gray snapper", "yellowtail snapper")
str_view_all(string = fish2, pattern = "(red|gray) snapper")
```


- Anchors
Two of the most common anchors are `^` & `$` which can be used to search for things at the start and end of a string respectively

```{r}
fish3 <-c("red snapper", "red grouper", "barred blenny", "gray snapper") 
grep(x = fish3, pattern="red", value = TRUE)
str_view_all(string = fish3, pattern="red")
str_view_all(string = fish3, pattern="^red")

My_files<- c("Fish_pictures.png", "data1.csv", "data2.csv", "merged_csvdata.xlsx")
str_view_all(My_files, ".csv")
str_view_all(My_files, "\\.csv$")
My_csvs<- My_files[str_detect(My_files, "\\.csv$")]
My_csvs
```
`list.files` function supports regular expressions
```{r eval=FALSE}
list.files(pattern="\\.csv$")
```

- Escaping  
Use `\\` to literally search for one of the special characters
```{r}
str_view_all(string = "This is a .", pattern = ".")
str_view_all(string = "This is a .", pattern = "\\.")
```


- Case Sensitive
Regular expressions are case sensitive. With base R functions, you can set `ignore.case = FALSE`. With stringr, you can put your pattern within a call to the `regex` function with `ignore_case=FALSE`  
```{r}
str_detect(string = "red grouper", pattern = "Grouper") #Case Sensitive
grepl(pattern = "Grouper", x = "red grouper", ignore.case = TRUE)
str_detect(string = tolower("red grouper"), pattern = tolower("Grouper"))
str_detect(string = toupper("red grouper"), pattern = toupper("Grouper"))
str_detect(string = "red grouper", pattern = regex("Grouper", ignore_case = TRUE))
```

- "all" functions  
Generally functions only act on the first match. If you are interested in all the matches, stringr has equivalent functions with "all" as a suffix. Base R also has equivalent functions.  
```{r}
test_string<- "abcd efg hijk"

str_replace(string = test_string, pattern = " ", replacement = "_")
sub(pattern = " ", replacement = "_", x = test_string)

str_replace_all(string = test_string, pattern = " ", replacement = "_")
gsub(pattern = " ", replacement = "_", x = test_string)
```


## Finish out our example together

Now that we've gone over some basics of regular expressions, let's finish our example by splitting the various components of column `site_code` into a `year`, `site`, and `sub_site` column:

```{r}
site_data
site_data <- site_data %>% mutate(year= str_extract(site_code, "^\\d{4}"))
head(site_data)
site_data <- site_data %>% mutate(site= str_extract(site_code, "[A-z ]+"))
head(site_data)
site_data <- site_data %>% mutate(sub_site= str_extract(site_code, "\\d$"))
site_data
```
By enabling perl syntax, you can also use lookarounds and conditionals  

For example we could use a Lookahead `?=` to say get the word in front of bay  
```{r}
str_view_all(site_data$site_code, regex("[A-z ]+(?= bay)", perl=TRUE))
```

## Summary

- Regular expressions are a crucial tool in the data analysis toolbox
- Regular expressions help us solve problems we may not be otherwise able to solve
- Regular expressions are supported in many functions in R

## Resources
https://rstudio.com/wp-content/uploads/2016/09/RegExCheatsheet.pdf  
http://edrub.in/CheatSheets/cheatSheetStringr.pdf  
https://stringr.tidyverse.org/articles/from-base.html
