---
title: "Connecting to Databases: SQL, SAS"
author: "M Schrandt"
date: "3/9/2021"
output: html_document
---

Most likely, your data are not stored in R, especially if there's a lot of it. Best data management practices indicate we should manage data in a relational database (e.g., SQL, MS Access, Oracle, SAS can be used like a relational database).

To access the data and take advantage of all R has to offer (wrangling, visualizing, analyzing), you could query from your database and save as .csv files and then load into R...or you can connect R directly to your database, which can aid in reproducibility.

Here's an RStudio resource for working with databases in R: https://db.rstudio.com/

If you connect to the database, you can query your data with R language/functions or, sometimes, you can use the native language (e.g., SQL).

I am by no means an expert at this, but I thought I'd share what has worked for me. My experience comes from the FIM data where we enter our data into a SQL Server database and then a copy of that SQL Server database is provided as part of a SAS library. We can connect directly to the SQL server database or we can read in the SAS library tables.

### 1. Connect to SQL Server Database and Query with R Functions

This option allows you to specify different connections, if you need to connect to different databases.

If you have datasets in the same SQL database but in different schema, the `dbplyr::in_schema()` function becomes very helpful when pulling/querying data.

This option allows you to filter/process data remotely before "collecting" it to your computer – if you can do your large query steps on the server, that might be preferred “collecting” data to your local machine. Keep in mind that in order to query data on the server, the R code has to be translated (in the background) to SQL language (this is what the `dbplyr` package can be used for), so not all functions are able to be translated.

Therefore, sometimes it can be beneficial to "collect" more to R and then process using R language if you're more comfortable in R.

I use a couple of very similar pacakges. There is probably a cleaner way to do this, but I use both, even though they provide similar funcionality (`odbc` and `DBI`). Here's a blog post: https://blog.revolutionanalytics.com/2017/08/a-modern-database-interface-for-r.html

The `odbc` package is built on top of the `DBI` package, which provides a low-level interface to databases 

A generalized connection for me looks like this:

```{r SQL_R_query, eval=FALSE}
#Load packages
library(odbc)    # has the odbc() function to use an ODBC connection to your server/data
library(DBI)     # has the dbConnect() function w/syntax dbConnect(drv, ...) where drv inhertis from DBIDriver
library(dbplyr)  # has the in_schema() function

#assign the database connection
db_connex <- dbConnect(odbc::odbc(),
                       driver = "SQL Server",
                       server = "[server name]",
                       database = "[database name]",
                       uid = "[user ID]",
                       pwd = "[password]")
```

To query, the `in_schema()` function has the syntax `in_schema(schema, table)` to help with tables.

```{r eval=FALSE}

#query a table from the database (example specific to FIM database)
timedat <- tbl(db_connex,in_schema("hsdb", "tbl_corp_physical_master")) %>%
  select(Reference, TripID, Gear, StartTime) %>%
  collect()  #use collect to actually pull the data down from SQL to your local R environment
```

The `dbplyr` package can transfer dplyr-like commands into SQL in the background. This works great if you'd like to query a bit more on the "server side" before "collecting" to your local R environment. Just be sure to pay attention to your functions and how they work or the arguments that they get. For example, the substring function in SQL has the syntax `SUBSTRING(string, start, length)`, which is  slightly different than some substring functions in R (base R `substr(x, start, stop)`.

Another thing to note is merging or joining. You can only join or merge datasets that are in the same space. For example, you can merge datasets that are in your R environment or you can merge ones still in the server environment, but you cannot merge a table in the server (that has not been "collected") with a table that is already in your R environment (or "collected").

### 2. Connect to SQL Server Database and use SQL code (copy & paste) in R

This option allows you to connect to the SQL database and then use exisiting SQL queries right from R. It uses the `dbGetQuery()` function.

This option is great if 1) you already have a collection of working SQL/Python code and you don't want to translate it into R and 2) you don't need to interact or change things in your exisitng SQL/Python code.

The `dbGetQuery()` function is best suited for `SELECT` queries, and not necessarily data manipulation.

```{r SQL_SQL_Query, eval=FALSE}
#Install & load packages
library(odbc)    # has the odbc() function to use an ODBC connection to your server/data
library(DBI)     # has the dbConnect() function
library(dbplyr)  # has the in_schema() function

#Connect to FIM SQL database
db_connex <- dbConnect(odbc::odbc(),
                       driver = "SQL Server",
                       server = "[server name]",
                       database = "[database name]",
                       uid = "[user ID]",
                       pwd = "[password]")

#Test the connection with a code to find a species name or NODCCODE from the species list table
Find.Species <- dbGetQuery(db_connex, "SELECT [NODCCODE]
                                             ,[TSN]
                                             ,[spp_code]
                                             ,[Scientificname]
                                             ,[Commonname]
                                             ,[Manufactured_nodccode]
                                             ,[Manufactured_tsn]
                                             ,[spp_code_old]
                                       FROM [FIMCorpInshore].[hsdb].[tbl_corp_ref_species_list]
                                       WHERE Commonname like '%pinfish%'
                                      ")
Find.Species
```

```Find.Species
    NODCCODE    TSN        spp_code      Scientificname       Commonname Manufactured_nodccode Manufactured_tsn  spp_code_old
1 8835430201 169187 Lag. rhomboides  Lagodon rhomboides          Pinfish                 FALSE            FALSE L. rhomboides
2 8835430401 169192 Dip. holbrookii Diplodus holbrookii Spottail Pinfish                 FALSE            FALSE  D. holbrooki
```

### 3. "Connect" to SAS database (library of SAS tables)

For SAS, the tables are typically housed in libraries, and although SAS itself is not a relational database, the tables can be set up in such a way that they are like a relational database.

In FIM, we mirror our SQL database as SAS data tables (because our staff originally queried and analyzed data with SAS and then graphed in SigmaPlot).

We can query the SAS data by simply reading in the SAS datasets. These are usually .sas7bdat files.

The idea is similar to a .csv file where you copy it locally, or point to the filepath, and then read it into R.

I use the `haven` package for this.

```{r SAS_Query, eval = FALSE}
# Load haven library (it's actually part of the tidyverse!)
library(haven)

# Read a SAS data table
read_sas(
  data_file,
  catalog_file = NULL,
  encoding = NULL,
  catalog_encoding = encoding,
  col_select = NULL,
  skip = 0L,
  n_max = Inf,
  cols_only = "DEPRECATED",
  .name_repair = "unique"
)
```

```{r SAS_Example, eval = TRUE}
library(haven)
path <- system.file("examples", "iris.sas7bdat", package = "haven")
read_sas(path)
```

### A look at FIM-specific examples...(demo)

### 4. MS Access

In FIM, our Gut Lab houses all their data in an MS Access database (not in SQL). There are also ways to connect to MS Access databases...

You'll likely need to install the appropriate driver from Microsoft. To check if you have the MS Access ODBC driver, you can run `odbcListDrivers()` in your console.

If "Microsoft Access Driver (.mdb, .accdb)" doesn't appear, you will need to install the Access redistributable ((Microsoft Access Database Engine 2016 Redistributable in some cases)

```{r MSAccess, eval=FALSE}
# full file path to Access DB
file_path <- "~/some_access_file.accdb"

# pass MS Access file path to connection string
accdb_con <- dbConnect(drv = odbc(), .connection_string = paste0("Driver={Microsoft Access Driver (*.mdb, *.accdb)};DBQ=",file_path,";"))


```

## FWRI R Club Announcements

<span style="color:green">Our next [remote/virtual] meeting via MS Teams is **April 13, 2021 at 2 pm eastern time. Anthony Gillis (FWC) will present on some spatial analyses in R**.</span>

I'm still looking for volunteers to present something at a meeting of your choice; please consider sharing your skills with the group. Or, if you're stuck on something and want feedback from the group, we can help!

Email if you're interested in presenting or if you have a topic you'd like someone else to present.

Thanks!
